XRONOS: an open data infrastructure for archaeological chronology

### 2024-11-29 paper.qmd
Add fig-ui

<!--
* [x] First draft
* [ ] Complete references:
  * [x] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
  * [ ] Appendix table
* [ ] Figures and tables
  * [x] FIGURE: graph of number of datasets through time
  * [x] FIGURE: map of number of dates in datasets
  * [ ] FIGURE: distribution of radiocarbon dates in c14bazAAR / IntChron / p3k14c / XRONOS
  * [x] FIGURE: data model
  * [ ] FIGURE: GUI screenshots (data browser, c14 show view, version history, data curation)
  * [ ] TABLE: list of 'issues'
  * [ ] APPENDIX: table of datasets
* [ ] Other TODOs
* [ ] Copyedit:
  * [ ] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
  -->

@@

Once ingested, we apply a number of automated and semi-automated quality control processes to integrate new data into the existing database.
Controlled vocabularies are used in a number of places in the data model (@FIG), and we use thesauri to automatically standardise these fields as much as possible.
For example, the taxonomic description of samples is controlled using GBIF's backbone taxonomy [@GBIF2023], and we also use a thesaurus service provided by GBIF to automatically change variant or obsolete taxonomic names to the canonical version.
If the system is not able to standardise a field using the available thesaurus, it is flagged for manual correction.
A wide variety of other potential data quality issues (e.g. missing data on what country a site is in) are also flagged for human review by this system [@tbl-issues], which can often be semi-automated (e.g. suggesting close matches in the thesaurus or the country indicated by the record's coordinates).

@@

<!-- TODO: figures? -->
<!-- TODO: add in more concrete examples -->

<!-- Public GUI -->
The graphical user interface (GUI) to XRONOS, accessed through a web browser (e.g. at <https://xronos.ch>), uses REST resources and actions as the building blocks for various interfaces through which users can browse, search, retrieve, and analyse chronometric data.
Each action on each resource is represented by a page, though not all of these are publicly accessible.
Pages representing REST resources directly are supplemented by a number of synthetic interfaces, for example the 'data browser' (<https://xronos.ch/data>), which facilitates more complex filtering, or the search interface (<https://xronos.ch/search>).
The GUI also includes several resources which are not part of XRONOS' scientific data model, for example documentation pages, user profiles and news articles; these are currently not exposed in the API.

>>>>

<!--
* [x] First draft
* [ ] Complete references:
  * [x] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
  * [ ] Appendix table
* [ ] Figures and tables
  * [x] FIGURE: graph of number of datasets through time
  * [x] FIGURE: map of number of dates in datasets
  * [ ] FIGURE: distribution of radiocarbon dates in c14bazAAR / IntChron / p3k14c / XRONOS
  * [x] FIGURE: data model
  * [x] FIGURE: GUI screenshots (data browser, c14 show view, version history, data curation)
  * [ ] TABLE: list of 'issues'
  * [ ] APPENDIX: table of datasets
* [ ] Other TODOs
* [ ] Copyedit:
  * [ ] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
  -->

@@

Once ingested, we apply a number of automated and semi-automated quality control processes to integrate new data into the existing database.
Controlled vocabularies are used in a number of places in the data model (@fig-data-model), and we use thesauri to automatically standardise these fields as much as possible.
For example, the taxonomic description of samples is controlled using GBIF's backbone taxonomy [@GBIF2023], and we also use a thesaurus service provided by GBIF to automatically change variant or obsolete taxonomic names to the canonical version.
If the system is not able to standardise a field using the available thesaurus, it is flagged for manual correction.
A wide variety of other potential data quality issues (e.g. missing data on what country a site is in) are also flagged for human review by this system [@tbl-issues], which can often be semi-automated (e.g. suggesting close matches in the thesaurus or the country indicated by the record's coordinates).

@@

::: {#fig-ui layout-nrow=2}
![Data browser](figures/xronos-ui-data-browser.png){#fig-ui-data-browser}

![Site record](figures/xronos-ui-show.png){#fig-ui-show}

![Record change log](figures/xronos-ui-papertrail.png){#fig-ui-papertrail}

![Curation interface](figures/xronos-ui-curation.png){#fig-ui-curation}

Views in the XRONOS web graphical user interface
:::

<!-- Public GUI -->
The graphical user interface (GUI) to XRONOS, accessed through a web browser (e.g. at <https://xronos.ch>), uses REST resources and actions as the building blocks for various interfaces through which users can browse, search, retrieve, and analyse chronometric data (@fig-ui).
Each action on each resource is represented by a page, though not all of these are publicly accessible.
The most import of these are the 'index' views, which list and summaries all instances of a resource and support filtering and sorting;
and 'show' views (e.g. @fig-ui-show for a site), which give a more comprehensive information on an individual record along with visualisations, links to related records, external linked open data resources, and a log of changes made to the data since it was imported into XRONOS (@fig-ui-papertrail).
Pages representing REST resources directly are supplemented by a number of synthetic interfaces, for example the 'data browser' (<https://xronos.ch/data>, @fig-ui-data-browser), which facilitates more complex filtering, or the search interface (<https://xronos.ch/search>).
The GUI also includes several resources which are not part of XRONOS' scientific data model, for example documentation pages, user profiles and news articles; these are currently not exposed in the API.
